---
title: "Analysis of Sample Means from the Exponential Distribution"
author: "Adam J"
date: '2020-05-27'
output: pdf_document
fig_crop: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8, fig.height = 4)
options(tinytex.verbose = TRUE)
library(ggplot2)
set.seed(5)
```

## Overview

This analysis is regarding investigating the central limit theorem, specifically with regards to the exponential distribution. 1000 Simulations of 40 exponentials are performed, where then the distribution of the sample mean and sample variance are compared to the population mean and variance. To highlight the normal property of the sample mean, as the number of samples grow large, the sample mean distribution is compared to that of the sampled exponential distribution. The analysis answer the following question: **Does the distribution of means of 40 exponentials behave as predicted by the Central Limit Theorem?** The Central Limit Theorem predicts that the distribution of averages of independent random variables follows that of the normal distribution. 

## Analysis
### Sample from the Exponential distribution

As a first step, 40000 observations are sampled from the exponential distribution with a rate of $\lambda=0.2$. In Figure 1, 1000 samples are plotted to show the distribution. According to the exponential distribution, the theoretical mean is $1/\lambda$, and the theoretical variance for a sample of size N is $1/N\lambda^2$.  

```{r, echo = FALSE}

# Number of simulations
sim <- 1000
# Number of exponentials
n <- 40
# Rate 
lambda <- 0.2 

# Creating observations
obs <- rexp(n*sim, rate = lambda)

# Store 1000 first observation to plot
subobs <- obs[1:1000]

# Plot distribution of 1000 exponentials
ggplot() +
  aes(subobs) + 
  geom_histogram(binwidth = 0.75, color = "black", fill = "steelblue") +
  geom_vline(xintercept = 5, color = "firebrick3", size = 1.5) + 
  labs(x = "Sample Value", y = "Counts") +
  labs(title = "1000 Samples from an Exponential Distribution with" ~ lambda ~ "= 0.2") +
  labs(caption = "Figure 1: Histogram of 1000 samples from 
  the exponential distribution with rate 0.2.") +
  coord_cartesian(xlim = c(0,45)) +
  theme(plot.title = element_text(size = 12)) +
  theme(plot.caption = element_text(hjust = 0.5))

```

### Sample Mean

The data is then separated into samples of 40 observations, which are then used to calculate 1000 sample means. The distribution of the sample means are plotted in Figure 2, together with the mean of the sample mean distribution in dashed green and the true mean of $1/\lambda=5$ in red. The density of the distribution is overlayed in order to more clearly visulize if it follows that of a normal distribution. 

```{r, echo = FALSE}

# Reformatting them into matrix form with dim [sim x n]
mobs <- matrix(obs, sim, n)

# Calculate sample means
sampleMeans <- apply(mobs, 1, mean)

# Calculate the mean of the sample means
muMean <- mean(sampleMeans)

# Calculate the theoretical mean
trueMean <- 1/lambda

ggplot() +
  aes(sampleMeans) + 
  geom_histogram(aes(y=..density..), binwidth = 0.25, color = "black", fill = "steelblue") +
  geom_density() +
  geom_vline(xintercept = muMean, color = "darkolivegreen3", lty = 5, size = 1.5) +
  geom_vline(xintercept = trueMean, color = "firebrick3", size = 1.5) + 
  labs(x = "Sample Mean", y = "Density") +
  labs(title = "Sample Means vs Theoretical Mean") +
  labs(caption = "Figure 2: Distribution of sample mean calculated on 1000 samples of 40 
  observations from the exponential distribution with rate 0.2.") +
  coord_cartesian(xlim = c(2,8)) +
  theme(plot.title = element_text(size = 12)) +
  theme(plot.caption = element_text(hjust = 0.5))

```

One can see that the distribution is centered around that of the theoretical mean. One can also clearly see the normal property of the distribution when when looking at the density and comparing to the exponential distribution in Figure 1.

\newpage
### Sample Variance 

Next the distribution of the sample variance is calculated and plotted in Figure 3, together with the mean of the sample variance distribution in dashed green and the theoretical variance of $1/N\lambda^2 = 0.625$ in red. 


```{r, echo = FALSE}

# Calculate sample variance
sampleVar <- apply(mobs, 1, var)/40

# Calculate the mean of the sample variance distribution
muVar <- mean(sampleVar)

# Calculate the theoretical variance for the sample
trueVar <- 1/(n*lambda^2)

ggplot() +
  aes(sampleVar) + 
  geom_histogram(binwidth = 0.1, color = "black", fill = "steelblue") +
  geom_vline(xintercept = muVar, color = "darkolivegreen3", lty = 5, size = 1.5) + 
  geom_vline(xintercept = trueVar, color = "firebrick3", size = 1.5) + 
  labs(x = "Sample Variance", y = "Counts") +
  labs(title = "Sample Variance vs Theoretical Variance") +
  labs(caption = "Figure 3: Histogram of sample variance calculated on 1000 samples of 40 
  observations from the exponential distribution with rate 0.2.") +
  coord_cartesian(xlim = c(0,3)) +
  theme(plot.title = element_text(size = 12)) + 
  theme(plot.caption = element_text(hjust = 0.5))

```

One can see that the mean of the sample variance line up well with the theoretical variance for 40 observations of exponentials.

## Conclusion

So to summarize and answer the question, **does the distribution of means of 40 exponentials behave as predicted by the Central Limit Theorem?** Yes, by plotting the distribution of the sample means and compare them to the normal distribution one can clearly see the expected normal property of the distribution.

\newpage
## Appendix

```{r ref.label=knitr::all_labels(), echo = T, eval = F}
```

