---
title: "Analysis of Relationship Between Vehicle Attributes and Fuel Efficiency"
author: "Adam J"
date: '2020-06-27'
output: pdf_document
fig_crop: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8)
library(tidyverse)
library(explore)
library(knitr)
```

## Overview
This study aims at investigating the relationship between vehicle attributes and fuel efficiency in miles per gallon (MPG), using the dataset `mtcars`. More specifically the study aims at answering the two following questions,

**Q1:** Is an automatic or manual transmission better for MPG?

**Q2:** Quantify the MPG difference between automatic and manual transmissions.

The outline of the study is as follows. First an exploration of the data is performed, after which model selection is performed by testing different candidates based on the results of the exploration. Residual plotting and diagnostic of the chosen model is then performed to ensure that no systematic errors are present. Uncertainty of the final model is quantified and presented, whereafter a conclusion to the two question of interest are presented.

### Exploratory Analysis 
In order to get an understanding of the relationship of MPG and transmission type, a boxplot is produced in Figure 1 which can be seen in the Appendix. One can clearly see that manual transmissions tend to have higher MPG than an automatic type. Let's investigate if this relationship hold true when accounting for other variables as well. In Figure 2, MPG is plotted against weight. One can observe that they share a strong relationship as well.

### Model selection
Let us first fit a model where we only consider transmission type.

```{r}
fit <- lm(mpg ~ am, data = mtcars)
summary(fit)$coefficients
```

We can see that the relationship is significant, with a p-value of `r summary(fit)$coefficients[2, 4]`, where the model estimate that fuel efficiency increase with `r summary(fit)$coefficients[2,1]` when using a manual transmission compared to an automatic. However we also observed a strong relationship between weight and fuel efficiency from the exploration. Let's fit a second model where we account for weight as well as transmission type.

```{r}
fit2 <- lm(mpg ~ am + wt, data = mtcars)
summary(fit2)$coefficients
```

Here we can see that transmission type is no longer significant, with a p-value of `r summary(fit2)$coefficients[2, 4]`. The confidence interval of the transmission regressor coefficient is [`r confint(fit2)[2,]` ], which contains zero. Weight is found to be a significant predictor of fuel efficiency instead with an estimate of `r summary(fit2)$coefficients[3,1]`, with a p-value of `r summary(fit2)$coefficients[3,4]` and a confidence interval of [`r confint(fit2)[3,]`]. Let us look at the residual plot to see if there are some variation left uncaptured. The plot is presented in the Appendix. One can see that the residuals are fairly normally distributed around zero. Let us also investigate the assumption of normality of the residuals directly by looking at a Q-Q plot, which can also be found in the Appendix. One can see that the residuals follows the line fairly well, which allow us to conclude that the residuals are normally distributed. 


## Conclusion
A multivariate regression was made to understand the relationship between transmission type and fuel efficiency in MPG. A first model where only transmission type was used as a regressor variable was fitted. The relationship between transmission type was found to be significant, where the regressor coefficient was positive with a value of `r summary(fit)$coefficients[2,1]` which indicates that a manual transmission is related to higher fuel efficiency. However when accounting for the weight of the vehicle, it was found that transmission type no longer was a significant predictor of MPG, with a confidence interval containing zero. Weight was found to have a significant negative relationship with fuel efficiency, which is expected. Diagnostic was performed on the latter model where no significant uncaptured variation was found. The residuals where also found be close to normally distributed, giving further confidence in the trustworthiness of our model that accounts for vehicle weight. With this information we can answer the questions posed. For question 1, there is not enough evidence to tell if one transmission type is better than the other, since no significant estimate on the relationship between transmission type and fuel efficiency could be found. For question 2, there is also not enough information to give a significant estimate of this quantity. 


\newpage
## Appendix

```{r, echo=FALSE}
g <- ggplot(mtcars, aes(x = as.factor(am), y = mpg, fill = as.factor(am)))
g +
  geom_boxplot() +
  labs(x = "Transmission type [0 - Automatic, 1 - Manual]", y = "Miles per Gallon") + 
  labs(title = "Fuel efficiency based on transmission type") +
  labs(caption = "Figure 1: Fuel efficiency in miles per gallon for vehicles by transmission type.") + 
  theme(plot.caption = element_text(hjust = 0.5, vjust = 0.5)) + 
  scale_fill_discrete(name = "Transmission", labels = c("Automatic","Manual"))
```



```{r, echo=FALSE}
g <- ggplot(mtcars, aes(x = wt, y = mpg))
g +
  geom_point(color = "steelblue") +
  labs(x = "Weight [1000 lbs]", y = "Miles per Gallon") + 
  labs(title = "Fuel efficiency compared to vehicle weight") +
  labs(caption = "Figure 2: Fuel efficiency in miles per gallon compared to vehicle weight.") + 
  theme(plot.caption = element_text(hjust = 0.5, vjust = 0.5))
```

```{r, echo=FALSE}
plot(fit2, which = c(1,2))
```





