---
title: "Human Activity Recognition Project"
author: "Adam J"
date: '2020-07-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Overview
Summary of the project. What is the purpose, what questions is of interest to answer, what was done in order to answer them? What was the main results and what conclusions can or can't we draw from them?

## Preprocessing
Clean the data 

 - Are there NA values? 
 - Is the data skewed? 

Create training, validation and test sets.

```{r}
trainDat <- read.csv("trainDat.csv")
testDat <- read.csv("testDat.csv")

# Select columns with no NA values
dat <- trainDat[, colSums(is.na(trainDat)) == 0]

```

## Data Exploration
Here we explore the data and look for pattern and think of potential candidates for model selection. 

 - Is there a nonlinear or linear relationship between the outcome (activity class) and the predictors (measurements from accelerometers)?
 - What is the approximate distribution of the variables in the dataset?

## Model selection
Investigate different models and how they compare on the validation set after having been trained on the training set.

## Model evaluation
Do a final evaluation on the selected model by testing to predict on unseen data from the test set.

## Discussion & Conclusion
Discuss the resulting accuracy or similar measures achieved in the final model evaluation and conclude if it is sufficient to answer the question of interest and if the purpose of the project was fulfilled.




