---
title: "Consequence analysis of weather events in US with respect to population health and economic cost based on observations from the NOAA database"
author: "Adam Josefsson"
date: '2020-04-30'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

The following study is an analysis of weather event obserations in US from the NOAA Storm Database, aiming to answer the two following questions:

**Q1: Which weather event is most harmful to population health?**

**Q2: Which weather event have the greatest economic consequences?**

Link to download the dataset: 
[Storm Data (47MB)](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)

Documentation for the dataset:
[National Weather Service Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

In this study the data is pre-processed, subsetted and then analysed, where total fatalities, total injuries, total property damage and total crop damage per weather event is calculated over the time period of January 1996 to November 2011 and plotted in two separate figures. Figure 1 tries to answer question 1 and includes information about total fatalities and total injuries per weather event. Figure 2 tries to answer question 2 and includes information about total property damages and total crop damages per weather event. The analysis is then concluded with a brief discussion.    

---

## Data processing

Start by loading the required packages, then if the data is not present on the local machine, download and load it. 

```{r, message=FALSE, warning=FALSE}
## Load required packages
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)

## Get data
if(!file.exists("repdata_data_StormData.csv.bz2")) {
  dataurl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  download.file(dataurl, "repdata_data_StormData.csv.bz2")
}

## Load data
dat <- tbl_df(fread("repdata_data_StormData.csv.bz2"))

```

After the data has been loaded, perform pre-processing by creating a new column of datetime objects from the BGN_DATE column. As a second step, let's explore the different types of event that are tracked over time. On the NOAA website it is mentioned ([link](https://www.ncdc.noaa.gov/stormevents/details.jsp)) that all types of events are tracked from 1996 and forward. Let's see for ourselves.

```{r}

# Transform date column to date object
dat <- mutate(dat, DATE =  mdy_hms(BGN_DATE))

# Let's now check how many different events that are being tracked after 1996
amount96 <- length(unique(filter(dat, DATE >= "1996-01-01")$EVTYPE))
print(paste0("Number of unique weather events from 1996 and forward: ", amount96))

```

From these results we can see that as much as `r amount96` unique types are being recorded from 1996 and forward. This seems to be a bit high. By inspecting the data, one can see that from 1993 there are a lot of different variants of recording the same type of event. In the documentation it is mentioned that the true number of unique types is infact 48. In order to extract the data of interest, the dataset is first filtered on dates from 1996 and forward. Filtering is then performed so that only observations with correct weather event label are used.


```{r}

dat <- filter(dat, DATE >= "1996-01-01")

# Offical weather events from NOAA database documentaion
evtypes <- c("Astronomical Low Tide", "Avalanche", "Blizzard", "Coastal Flood",
            "Cold/Wind Chill", "Debris Flow", "Dense Fog", "Dense Smoke", 
            "Drought", "Dust Devil", "Dust Storm", "Excessive Heat",
            "Extreme Cold/Wind Chill", "Flash Flood", "Flood", "Frost/Freeze",
            "Funnel Cloud", "Freezing Fog", "Hail", "Heat",
            "Heavy Rain", "Heavy Snow", "High Surf", "High Wind",
            "Typhoon",  "Ice Storm", "Lake-Effect Snow", "Lakeshore Flood",
            "Lightning", "Marine Hail", "Marine High Wind", "Marine Strong Wind",
            "Marine Thunderstorm Wind", "Rip Current", "Seiche", "Sleet",
            "Storm Surge/Tide", "Strong Wind", "Thunderstorm Wind", "Tornado", 
            "Tropical Depression", "Tropical Storm", "Tsunami", "Volcanic Ash", 
            "Waterspout", "Wildfire", "Winter Storm", "Winter Weather")

dat <- filter(dat, EVTYPE %in% toupper(evtypes))

```

Filtering is then performed to create two new subsets of the data to be used for the two questions respectively. The first subset is called *harmdat*, containing information about the harmfulness of events, and is created by filtering out all events which have either non-zero fatalities or non-zero injuries.

```{r}


# Filter out data with events which have non-zero recorded values for fatalities or injuries
harmdat <- filter(dat, (FATALITIES != 0) | (INJURIES != 0))


```

The second subset of the data is called *econdat*, and contains all events with non-zero recorded values for property or crop damage. To enable us to take this subset, we must first create a new column merging the information in the PROPDMG and CROPDMG variables with the PROPDMGEXP and CROPDMGEXP variables, respectively. The dataset is then filtered on observations which have either non-zero property damages or non-zero crop damages. 

```{r}

# Create vector with multiper values for each present symbol in EXP variables in the filtered dataset
multipliers <- c("0"=10, "K"=1e3, "M"=1e6, "B"=1e9)

# Create column in which information from propdmg and propdmgexp is combined
econdat <- mutate(dat, ACTPROPDMG = if_else(PROPDMGEXP != "", PROPDMG * multipliers[PROPDMGEXP], 0))

# Create column in which information from cropdmg and cropdmgexp is combined
econdat <- mutate(econdat, ACTCROPDMG = if_else(CROPDMGEXP != "", CROPDMG * multipliers[CROPDMGEXP], 0))

# Filter out observations which have either non-zero property damages or non-zero crop damages 
econdat <- filter(econdat, (ACTPROPDMG != 0 | ACTCROPDMG != 0))

```

The data is now ready for analysis! 

---

## Results

In this section we perform the analysis for the two questions and present the results, with one subsection per question. The first subsection tries to present result to answer question 1, which event is most harmhul to the population, and the second subsection tries to present result to answer question 2, which event is causing the most economic damage.  

### Analysis for Q1

In order to answer the first question regarding which weather event that is most harmful for the population, the data is grouped by weather event and then both total fatalities and total injuries over time are calculated, using the pre-processed dataset *harmdat*. One could also argue that one should look at the mean fatalities and injuries per event as that would tell how deadly or harmful one individual event would be on average. However now we take the perspective of which event has caused the most harm to the population over time, i.e. very deadly but rare events are not considered as more harmful compared to an event which is common with a low-to-medium deadliness which is causing a lot of harm over time.

```{r}

# Calculate total fatalities and total injuries per weather event
totharmdat <- harmdat %>% group_by(EVTYPE) %>% summarize(total_fatalities = sum(FATALITIES), total_injuries = sum(INJURIES))

# Print the top 10 weather events ranked by total fatalities
totharmdat %>% arrange(desc(total_fatalities)) %>% top_n(n = 10, wt = total_fatalities)

```

From these result one can see that when ranking the weather events by only total fatalities, excessive heat is ranked to be the most harmful. However one can observe that tornadoes, which is ranked number two in deadliness, has relatively high amount of caused injuries. In order to get a more clear view of how much harm a specific event is causing the population, total fatalities and total injuries are summed together and the top ten most harmful events are filtered out and plotted in Figure 1.

```{r, fig.width=10, fig.height=6, dpi=300}

# Calcultate total harm, filter out the top ten most harmful events and transform the dataset to long format (this will simplify plotting)
top10harmdat <- totharmdat %>% 
                group_by(EVTYPE) %>%
                mutate(total_harm = total_fatalities + total_injuries) %>%
                arrange(desc(total_harm)) %>%
                .[1:10,] %>%
                pivot_longer(c(-EVTYPE,-total_harm))

# Plot
top10harmdat %>% 
                ggplot(aes(x = reorder(EVTYPE, total_harm), y = value, fill = name)) + 
                geom_bar(stat = "identity") + 
                labs(x = "Weather event", y = "Count") +
                labs(title = "Harm Caused to Population by Weather Events in US 1996-2011") +
                labs(subtitle = "Number of affected persons and outcome") +
                labs(caption = "Figure 1: The ten weather event which has caused the most harm to the population of US \nin terms of number of affected persons and outcome, between January 1996 and November 2011.") +
                theme(plot.caption = element_text(hjust = 0.5, vjust = -1)) +
                theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
                theme(plot.margin = unit(c(1,0,1,0.2),"cm")) + 
                scale_fill_discrete(name = "Outcomes", labels = c("Fatalities", "Injuries"))
                


```

One can observe more clearly the magnitude of the difference between harmfulness in events, where tornadoes far exceeds the other weather types. Even though excessive heat have caused sligthly more fatalities, the number of injuries caused by tornadoes is a factor of three larger than the injuries caused by excesive heat. 

### Analysis for Q2

In this section we perform analysis and present results to answer the second question, which weather event has most severe economic consequences. In order to perform the analysis, we start from the pre-processed data *econdat*. As a first step in the analysis we group the data per weather event and calculate the total property damages and total crop damages and print out the top ten weather events ranked by total property damages. 

```{r}

# Calculate total property damage and total crop damage per weather event
totecondat <- econdat %>% group_by(EVTYPE) %>% summarize(total_propdmg = sum(ACTPROPDMG), total_cropdmg = sum(ACTCROPDMG))

# Print the top 10 weather events ranked by total property damage 
totecondat %>% arrange(desc(total_propdmg), desc(total_cropdmg)) %>% top_n(n = 10, wt = total_propdmg)

```

From these results one can observe that floods have a value of total property damages with a magnitude higher than the second highest, tornado. Let's observe these results in a plot, where we get a clearer view of the extent of crop damages as well. This can be seen in Figure 2.

```{r, fig.width=10, fig.height=6, dpi=300}

# Calcultate total harm, filter out the top ten most harmful events and transform the dataset to long format (this will simplify plotting)
top10econdat <- totecondat %>% 
                group_by(EVTYPE) %>%
                mutate(total_cost = total_propdmg + total_cropdmg) %>%
                arrange(desc(total_cost)) %>%
                .[1:10,] %>%
                pivot_longer(c(-EVTYPE,-total_cost))

# Plot
top10econdat %>% 
                ggplot(aes(x = reorder(EVTYPE, total_cost), y = value/10^9, fill = name)) + 
                geom_bar(stat = "identity") +
                labs(x = "Weather event", y = "Cost [Billion USD]") +
                labs(title = "Economic Consequences from Weather Events in US 1996-2011") +
                labs(subtitle = "Cost of damages in billion USD and type of damage") +
                labs(caption = "Figure 2: The ten weather events with the highest total caused economic \ndamage in the US between January 1996 and November 2011.") +
                theme(plot.caption = element_text(hjust = 0.5, vjust = -1)) +
                theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
                theme(plot.margin = unit(c(1,0,1,0.2),"cm")) + 
                scale_fill_discrete(name = "Type", labels = c("Crop damage", "Property damage"))

```

One can again see that floods vastly outnumber any other weather event in terms of economic consequences. A potential explanation for this can be found in the documentation, where it is mentioned that floods is the only event where the reporter is *required* to enter monetary damages even if it is a "guesstimate". This could obviously skew the data. Another observation one can make in the plot is that the event which is causing the highest crop damage is drought, which could be expected.

---

## Discussion and conclusion

We can now conclude the analysis and answer the questions. The answer to **question 1**, which event is most harmful to the population, is clearly **tornadoes** with over 20000 affected persons between 1996 and 2011. The answer to **question 2**, which event has greatest economic consequneces, is by a great margin **floods**. It could however be argued that this result is partly skewed due to mandatory filling in of monetary damages for floods even if they are inaccurate. For other events where the monetary damage is unclear, the value might be left blank in the report, further skewing the data. 


---